---
title: "Correlation of Predicted KEGG Pathways and OTUs"
author: "Hui/Jonathan"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
setwd("B:/")
source("fcns/config.r")
setwd("external/altindis/celiac_disease_16sn/corr")
library(pheatmap)
```

## Data
OTU data [otu.rda](../limma/otu.rda),  
Differential OTU statistcs [otu_stats.csv](../limma/otu_stats.csv).  

PICRUSt2 KEGG pathway data [kegg_pwys.rda](../picrust2/kegg_pwys.rda),  
Differential KEGG pathway statistcs [kegg_pwys_stats.csv](../picrust2/kegg_pwys_stats.csv).  

```{r parse}
load("../limma/otu.rda")
annot_otu <- annot; rm(annot)

load("../picrust2/kegg_pwys.rda")
annot_kegg <- annot; rm(annot, md)
stopifnot(colnames(logcpm) == colnames(pwys.tab))

dat_otu <- read.csv("../limma/otu_stats.csv", row.names = 1)
dat_kegg <- read.csv("../picrust2/kegg_pwys_stats.csv", row.names = 1)
```

## Correlation
```{r corr}
ntop <- 20
ages <- c(1, 2.5, 5)
corr_res <- list()
for(age in ages){
  pcol <- paste0("Celiac_vs_Control_in_Presort_Age", age, ".p")
  topo <- rownames(dat_otu)[order(dat_otu[, pcol])][1:ntop]
  topk <- rownames(dat_kegg)[order(dat_kegg[, pcol])][1:ntop]
  samp <- rownames(pheno)[pheno$Sort == "Presort" & pheno$Age == paste0(age, "year")]
  
  corr_tab <- matrix(0, nrow = ntop, ncol = ntop, dimnames = list(topo, topk))
  for(o in topo){
    for(k in topk){
      corr_tab[o, k] <- cor(x = logcpm[o, samp], y = pwys.tab[k, samp], method = "pearson")
    }
  }
  dimnames(corr_tab) <- list(annot_otu[topo, "Taxa"], annot_kegg[topk, "Pathway"])
  
  nm <- paste0("Age", age, "_presort")
  corr_res[[nm]] <- data.frame(Taxa = rownames(corr_tab), corr_tab, check.names = FALSE)
  
  num <- max(nchar(rownames(corr_tab)))
  pheatmap(corr_tab, cluster_rows = TRUE, cluster_cols = TRUE, color = colorRampPalette(c("blue", "white", "red"))(100), main = "Pearson Correlation",
           filename = paste0(nm, "_otu_kegg_corr_heat.pdf"), height = 8, width = 8 + num/25)
}

writexl::write_xlsx(corr_res, "presort_otu_kegg_corr.xlsx")
```

We performed Pearson correlation tests for the selected metabolites and predicted KEGG pathways.  The heatmaps [Age1_presort_otu_kegg_corr_heat.pdf](./Age1_presort_otu_kegg_corr_heat.pdf), [Age2.5_presort_otu_kegg_corr_heat.pdf](./Age2.5_presort_otu_kegg_corr_heat.pdf), and [Age5_presort_otu_kegg_corr_heat.pdf](./Age5_presort_otu_kegg_corr_heat.pdf) show the correlation coefficients.  
