---
title: "Differential OTUs"
author: "Hui/Jonathan"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: word_document
bibliography: B:/annotations/bib/bioinfo.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
setwd("B:/")
source("fcns/config.r")
source("fcns/select_top.r")
source("fcns/fcns_limma/edger_contrasts.r")
setwd("external/altindis/celiac_disease_16sn/limma")
library(plyr)
library(jdcbioinfo)
```

## Purpose
Differential OTUs by using limma [@ritchie_2015].  
Results are in folder [limma](../limma).

## Data
OTU table [otu_table.csv](../phyloseq/otu_table.csv), 
Taxonomy table [taxonomy_table.csv](../phyloseq/taxonomy_table.csv), 
Metadata table [metadata_table.csv](../phyloseq/metadata_table.csv)

```{r parse}
counts <- data.matrix(read.csv("../phyloseq/otu_table.csv", row.names = 1,  skipNul = TRUE))
pheno <- read.csv("../phyloseq/metadata_table.csv", row.names = 1,  skipNul = TRUE)
annot <- read.csv("../phyloseq/taxonomy_table.csv", row.names = 1,  skipNul = TRUE)

annot$Taxa <- apply(annot, 1, function(v) paste(v[!is.na(v)], collapse = "_"))
pheno$Disease.status <- factor(pheno$Disease.status, levels = c("Control", "Celiac"))
pheno$Group <- factor(pheno$Group, levels = unique(pheno$Group))
```

We combine the counts of each two technical replicates.   

## Filtering
```{r filt}
dge <- DGEList(counts = counts)
dge <- dge[rowSums(counts >= 10) > 5, ]
dim(dge)

# donot using  calcNormFactors()

des <- model.matrix(~0+Group, pheno)
colnames(des) <- gsub("Group", "", colnames(des))
el <- voom(dge, design = des, plot = TRUE, span = 0.5)
logcpm <- el$E
rm(dge, el)
save(logcpm, annot, pheno, file = "otu.rda")
```

To filter out low abundant OTUs, we keep OTUs that have read counts at least 10 in 5 samples. There are 611 OTUs after filtering. 

## Test for differential OUTs
```{r limma}
contr.v <- c(Celiac_vs_Control_in_Presort_Age1 = "Presort_Celiac_1 - Presort_Control_1",
             Celiac_vs_Control_in_IGneg_Age1 = "IGneg_Celiac_1 - IGneg_Control_1",
             Celiac_vs_Control_in_IGpos_Age1 = "IGpos_Celiac_1 - IGpos_Control_1", 
             IGpos_vs_IGneg_in_Control_Age1 = "IGpos_Control_1 - IGneg_Control_1",  
             IGpos_vs_IGneg_in_Celiac_Age1 = "IGpos_Celiac_1 - IGneg_Celiac_1",
             
             Celiac_vs_Control_in_Presort_Age2.5 = "Presort_Celiac_2.5 - Presort_Control_2.5",
             Celiac_vs_Control_in_IGneg_Age2.5 = "IGneg_Celiac_2.5 - IGneg_Control_2.5",
             Celiac_vs_Control_in_IGpos_Age2.5 = "IGpos_Celiac_2.5 - IGpos_Control_2.5", 
             IGpos_vs_IGneg_in_Control_Age2.5 = "IGpos_Control_2.5 - IGneg_Control_2.5",  
             IGpos_vs_IGneg_in_Celiac_Age2.5 = "IGpos_Celiac_2.5 - IGneg_Celiac_2.5",
             
             Celiac_vs_Control_in_Presort_Age5 = "Presort_Celiac_5 - Presort_Control_5",
             Celiac_vs_Control_in_IGneg_Age5 = "IGneg_Celiac_5 - IGneg_Control_5",
             Celiac_vs_Control_in_IGpos_Age5 = "IGpos_Celiac_5 - IGpos_Control_5", 
             IGpos_vs_IGneg_in_Control_Age5 = "IGpos_Control_5 - IGneg_Control_5",  
             IGpos_vs_IGneg_in_Celiac_Age5 = "IGpos_Celiac_5 - IGneg_Celiac_5")
mtt <- limma_contrasts(logcpm, grp = pheno$Group, contrast.v = contr.v, trend = TRUE)
mtt.df <- data.frame(signif(mtt, 3), annot[rownames(mtt), ])

signif_hist(mtt, name = "signif_hist", pi0 = TRUE)
write.csv(mtt.df, "otu_stats.csv", na = "")
```

To discover the differential OTUs, we use limma, an R package that powers differential expression analyses [@ritchie_2015]. We perform moderated t-tests for the following comparisons: between any 2 disease status at different age and sorting, or between IGpos and IGneg in different age and disease status.    
The histograms of significance are at [signif_hist.pdf](./signif_hist.pdf). If no OTUs were associated with the phenotype, we would expect the p-value histogram to be flat and all FDRs to be near one. The more associated OTUs there are, the more enrichment there is at low p-values, the lower will be the FDRs.  We also estimate the proportion of the true null hypothesis (i.e. non-significant OTUs) [@langaas_2005].    
OTU statistics tables for all OTUs [otu_stats.csv](./otu_stats.csv). The table contains the average logCPM of each group, p-values, FDR, log fold-change, fold-change, and taxonomy information.      

## Plots
```{r plots}
## boxplot
topotu1 <- select.top(mtt, contrasts.v = contr.v[grep("Presort", names(contr.v))])
pdf("top_otus_boxplot_presort.pdf", 9, 3)
for (otu in topotu1){
    dat2p <- data.frame(logCPM = logcpm[otu, pheno$Sort == "Presort"], pheno[pheno$Sort == "Presort", ])
     
    ggp <- ggplot(data = dat2p, mapping = aes(x = Disease.status, y = logCPM)) + theme_bw()
    ggp <- ggp + ggtitle(annot[otu, "Taxa"]) + theme(legend.position = "none")
    ggp <- ggp + geom_boxplot(mapping = aes(fill = Disease.status)) + facet_grid(~ Age)
    plot(ggp)
}
dev.off()

topotu2 <- select.top(mtt, contrasts.v = contr.v[-grep("Presort", names(contr.v))])
pdf("top_otus_boxplot_postsort.pdf", 9, 5)
for (otu in topotu2){
    dat2p <- data.frame(logCPM = logcpm[otu, pheno$Sort != "Presort"], pheno[pheno$Sort != "Presort", ])
     
    ggp <- ggplot(data = dat2p, mapping = aes(x = Disease.status, y = logCPM)) + theme_bw()
    ggp <- ggp + ggtitle(annot[otu, "Taxa"]) + theme(legend.position = "none")
    ggp <- ggp + geom_boxplot(mapping = aes(fill = Disease.status)) + facet_grid(Sort~ Age)
    plot(ggp)
}
dev.off()

## heatmaps
pheno.df1 <- pheno[pheno$Sort == "Presort", c("Disease.status", "Age")]
gaps.col1 <- which(diff(as.numeric(as.factor(pheno.df1$Age)), lag=1) != 0)
ezheat(logcpm[topotu1, pheno$Sort == "Presort"], pheno.df = pheno.df1,  sc = "z", reorder_rows = TRUE, labrows = annot[topotu1, "Taxa"], labcols = "",
       gaps_col = gaps.col1, main = "logCPM", name = "top_otus_heat_presort", height = 8, width = 16, clip = 2)


pheno.df2 <- pheno[pheno$Sort != "Presort", c("Disease.status", "Age", "Sort")]
gaps.col2 <- which(diff(as.numeric(as.factor(pheno.df2$Age)), lag=1) != 0)
ezheat(logcpm[topotu2, pheno$Sort != "Presort"], pheno.df = pheno.df2,  sc = "z", reorder_rows = TRUE, labrows = annot[topotu2, "Taxa"], labcols = "",
       gaps_col = gaps.col2, main = "logCPM", name = "top_otus_heat_postsort", height = 8, width = 25, clip = 2)
```

Equal numbers of top OTUs (based on p-values) are selected from each comparison. The boxplots for top OUTs are at [top_otus_boxplot_presort.pdf](./top_otus_boxplot_presort.pdf), and [top_otus_boxplot_postsort.pdf](./top_otus_boxplot_postsort.pdf). The same sets of top OTUs are use in heatmaps [top_otus_heat_presort.pdf](./top_otus_heat_presort.pdf) and [top_otus_heat_postsort.pdf](./top_otus_heat_postsort.pdf).


```{r check}
stopifnot(rownames(counts) == rownames(annot))
stopifnot(colnames(counts) == rownames(pheno))
stopifnot(mtt.df[1, "Taxa"] == "Bacteria_Proteobacteria_Alphaproteobacteria_Rhizobiales_Brucellaceae_Brucella")
```

## Reference
  